\bf Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-02 paper.qmd
liver example added

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tourr)
library(mulgar)
library(dplyr)
library(ggplot2)
library(patchwork)
```

@@

### Medical patients

>>>>

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tourr)
library(mulgar)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
```

@@

### Health: liver function tests

This example is motivated by a problem posed during consulting with a pharmaceutical company, but the data shown here is simulated, simply to illustrate the application. Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT). There are normal ranges on these measurements reported by @lib-med-liver-norms and listed in @tbl-liver-norms.

```{r}
#| label: tbl-liver-norms
#| tbl-cap: "Normal ranges provided for liver function tests."
liver_stats <- tribble(
  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
  5,     5,   30,      35,        0,    2, 60,   "m", "min",
  40,     30,  120,      50,        20,   44, 80,   "m", "max",
  5,     5,   30,      35,        0,    2, 60,   "f", "min",
  40,     30,  120,      50,        20,   70, 80,    "f", "max"
)
liver_stats |> 
  filter(sex == "m") |>
  select(-sex) |>
  select(stat, albumin, protein, bilirubin, GGT, AST, ALP, ALT) |>
  kable(
    row.names = FALSE,
    col.names = c("", "albumin (g/L)", "protein (g/L)", 
        "bilirubin (µmol/L)", "GGT (IU/L)", "AST (IU/L)", 
        "ALP (IU/L)", "ALT (IU/L)"),
    booktabs = TRUE
  )
```

These measurements are also likely correlated, based on guidance like the *ratio of AST to ALT of 2:1 indicates possible alcohol abuse*. Although this was provided by the pharmaceutical company, correlation between these measurements for normal patients is not readily available. When a correlation matrix for normal patients is provided this would allow construction of the null ellipse upon which to examine new samples.

```{r}
#| eval: false
#| echo: false
liver_stats <- tribble(
  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
)
set.seed(1158)
norm_vc <- diag(1, 4, 4)
norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
for (i in 1:3)
  for (j in (i+1):4)
    norm_vc[j, i] <- norm_vc[i, j]
colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
set.seed(217)

women_vc <- norm_vc
women_vc[1,4] <- 0.5
women_vc[4,1] <- 0.5
women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
                 sigma = women_vc)
colnames(women) <- c("GGT", "AST", "ALP", "ALT")
women <- as.data.frame(women)

animate_xy(women,            
           ellipse=norm_vc, 
           axes = "bottomleft", 
           half_range=5, 
           center=FALSE)

animate_xy(women, guided_anomaly_tour(anomaly_index(),
                                    ellipse=norm_vc), 
           ellipse=norm_vc, 
           axes = "bottomleft", 
           half_range=5, 
           center=FALSE)

set.seed(1208)
pt1 <- tibble(GGT = rnorm(5, mean=1.3, sd=0.15),
              AST = rnorm(5, mean=1, sd=0.2),
              ALP = c(1, 1.3, 1.6, 2.8, 2.9), 
              ALT = c(-1, -1.3, -1.6, -2.9, -3.1))
animate_xy(pt1, guided_anomaly_tour(anomaly_index(),
                                     ellipse=norm_vc), 
           ellipse=norm_vc, 
           axes = "bottomleft", 
           half_range=5, 
           center=FALSE,
           edges = matrix(c(1:4, 2:5), byrow=FALSE, ncol=2),
           obs_labels = c("45", "50", "55", "65", "70"))
```

@fig-liver illustrates two examples. The first is similar to the consulting project. A sample of liver test scores for new patients was provided in order to examine their values relative to the normal range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows a projection of this sample relative to the normal range. Three of the patients are outside the confidence ellipse but all of the patients are located away from the mean. What has been typical in the past is to compute normal values based on tests of healthy young males. The samples provided for the project were all recorded on women. We see that this sample has slightly lower ALT and ALP, which is consistent with what is reported in @lib-med-liver-norms.

The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70. The lines connect the records in tome order. The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:

$$
P = \left[ \begin{array}{rr}
            0.371 & -0.128 \\
            -0.388 & 0.732 \\
            0.140 & 0.599 \\
            -0.832 & -0.297 
            \end{array} \right]
$$

From the axes representation in @fig-liver (b) of this projection, we can see that the direction that profile extends is primarily contrasting ALT (fourth row) and ALP (third row). This is consistent with aging, where ALP increases and ALT decreases.

::: {#fig-liver layout-ncol=2}

![Sample of female patients.](images/liver1.png){#fig-liver1 width=300 align="center"}

![Longitudinal profile of an aging patient.](images/liver2.png){#fig-liver2 width=300 align="center"}

Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient. Red cross indicates observation is outide the 4-D confidence ellipse. The female patients tend to have lower ALP and ALT than the normal range. As the patient aged, the level of ALP increases and ALT decreases.
:::

### 2024-07-02 paper.tex
liver example added

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

The projection of an ellipsoid onto \gD{} is an ellipse, where the curve
of the ellipse is defined through the set of points \(\mathbfit{x}\) for
which the gradient is parallel to the projection plane. That is, the
curve consists of \(\mathbfit{x}\) satisfying

\[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
\]

for some \(\mathbfit{s} \in \mathbb{R}^2\), where \(P\) is a
\((p\times 2)\) orthonormal basis defining the projection. We can write
\(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\). Making this
substitution in the \pD{} ellipsoid equation yields

\[c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
\]

We call points in the projection that are on the curve \(\mathbfit{y}\).
From this we can compute the analog equation for the projection as

\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.\]
This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
the \gD{} projection. In general \(c\) could be any constant, but
typically we would select it as a quantile of the \(\chi^2\)
distribution, so that the size of the ellipse corresponds to a selected
probability.

@@

\subsection{Medical patients}\label{medical-patients}

\subsection{Robust statistics}\label{robust-statistics}

\subsection{Physics?}\label{physics}

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newtheorem*{theorem}{Theorem}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

\begin{theorem}
  The projection of this \pD{} ellipsoid in \gD{} has the equation
$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.$$
\end{theorem}
\begin{proof}
The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane. That is, the curve 
consists of $\mathbfit{x}$ satisfying

$$\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
$$

for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
Making this substitution in the \pD{} ellipsoid equation yields

$$c^2 = (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T
$$

We call points in the projection that are on the curve $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for $\mathbfit{x}$ on the \pD{} curve. Then $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, where $\mathbfit{\mu}_p = \mathbfit{\mu} P$ is the projected mean. We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in the equation \(c^2 = \mathbfit{s} P^T \Sigma P \mathbfit{s}^T\). From this we can compute the analog equation for the projection as

$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$

as claimed.
\end{proof}

This means the matrix \((P^T \Sigma P)^{-1}\) is defining the ellipse in
the \gD{} projection. In general \(c\) could be any constant, but
typically we would select it as a quantile of the \(\chi^2\)
distribution, so that the size of the ellipse corresponds to a selected
probability.

@@

\subsection{Health: liver function
tests}\label{health-liver-function-tests}

This example is motivated by a problem posed during consulting with a
pharmaceutical company, but the data shown here is simulated, simply to
illustrate the application. Liver function tests commonly provide
measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase
(GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and
alanine aminotransferase (ALT). There are normal ranges on these
measurements reported by \citet{lib-med-liver-norms} and listed in
Table~\ref{tbl-liver-norms}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0421}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1474}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2000}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1158}}@{}}

\caption{\label{tbl-liver-norms}Normal ranges provided for liver
function tests.}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
albumin (g/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
protein (g/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
bilirubin (µmol/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
GGT (IU/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
AST (IU/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ALP (IU/L)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ALT (IU/L)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
min & 35 & 60 & 0 & 2 & 5 & 30 & 5 \\
max & 50 & 80 & 20 & 44 & 30 & 120 & 40 \\

\end{longtable}

These measurements are also likely correlated, based on guidance like
the \emph{ratio of AST to ALT of 2:1 indicates possible alcohol abuse}.
Although this was provided by the pharmaceutical company, correlation
between these measurements for normal patients is not readily available.
When a correlation matrix for normal patients is provided this would
allow construction of the null ellipse upon which to examine new
samples.

Figure~\ref{fig-liver} illustrates two examples. The first is similar to
the consulting project. A sample of liver test scores for new patients
was provided in order to examine their values relative to the normal
range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows
a projection of this sample relative to the normal range. Three of the
patients are outside the confidence ellipse but all of the patients are
located away from the mean. What has been typical in the past is to
compute normal values based on tests of healthy young males. The samples
provided for the project were all recorded on women. We see that this
sample has slightly lower ALT and ALP, which is consistent with what is
reported in \citet{lib-med-liver-norms}.

The second example shows a longitudinal record of a single patient,
measured at ages 45, 50, 55, 65 and 70. The lines connect the records in
tome order. The projection corresponds to the maxima from a
projection-pursuit guided tour using the anomaly index:

\[
P = \left[ \begin{array}{rr}
            0.371 & -0.128 \\
            -0.388 & 0.732 \\
            0.140 & 0.599 \\
            -0.832 & -0.297 
            \end{array} \right]
\]

From the axes representation in Figure~\ref{fig-liver} (b) of this
projection, we can see that the direction that profile extends is
primarily contrasting ALT (fourth row) and ALP (third row). This is
consistent with aging, where ALP increases and ALT decreases.

\includegraphics[width=3.125in,height=\textheight]{images/liver1.png}

\subcaption{\label{fig-liver1}Sample of female patients.}

\includegraphics[width=3.125in,height=\textheight]{images/liver2.png}

\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient.}

\caption{\label{fig-liver}Two projections of simulated example data
corresponding to common liver tests: (a) sample of female patients, (b)
longitudinal test results for a single patient. Red cross indicates
observation is outide the 4-D confidence ellipse. The female patients
tend to have lower ALP and ALT than the normal range. As the patient
aged, the level of ALP increases and ALT decreases.}

\subsection{Robust statistics: weather
extremes}\label{robust-statistics-weather-extremes}
